MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS

### 2025-01-01 review/response_241203.tex

\end{editor}

\begin{reviewer}
	\summary{
		This manuscript investigates the impact of \gls{bd}-\gls{ris} to the distribution of singular values of \gls{mimo} channel of a point-to-point communication system. Especially, the authors derive bounds of singular values of \gls{bd}-\gls{ris} channel and propose algorithms to design \gls{bd}-\gls{ris} configuration.
	}

	\comment{
		The contribution of Sec. \ref{sc:framework} is unclear for the following two reasons.
		\begin{enumerate}
			\item The manuscript claims to propose a universal optimization framework for \gls{bd}-\gls{ris} as in \eqref{op:general}. However, the framework in \eqref{op:general} is indeed not sufficiently general. It does not consider \gls{bd}-\gls{ris} dependent constraints (e.g., \gls{qos}, sensing quality) besides the orthogonal conditions \eqref{cn:general_unitary}. The reviewer notice that a more general framework, which considers \gls{bd}-\gls{ris} constraints in addition to \eqref{cn:general_unitary}, has already been investigated and solved in the authors' previous work \cite{Zhou2023}.
			\item Besides, both the non-geodesic \gls{rcg} (Sec. \ref{sc:rcg_general}) and geodesic \gls{rcg} (Sec. \ref{sc:rcg_geodesic}) algorithms, e.g., \eqref{eq:geodesic_iden}--\eqref{eq:iterate_mul}, are well established methods in the literature, e.g., \cite{Abrudan2008}. Therefore, the novelty of Sec. \ref{sc:framework} unclear.
		\end{enumerate}
	}

	\reply{
		We appreciate the reviewer's insightful comments. We have revised the manuscript to clarify the contributions of Sec. \ref{sc:framework} and the novelty of the proposed optimization framework. Specifically, we have added a discussion on the limitations of the proposed framework in \eqref{op:general} and the novelty of the proposed geodesic \gls{rcg} method. We have also added a comparison with the existing literature to highlight the novelty of the proposed framework. We believe that the revised manuscript now clearly articulates the contributions of Sec. \ref{sc:framework} and the novelty of the proposed optimization framework.
	}

@@

	\comment{
		In Sec. \ref{sc:pareto}, the manuscript proposes to maximize weighted sum of singular values of the effective channel $\mathbf{H}$. The optimization objective is rather vague. Is the weighted sum of singular values related to any meaningful performance metric (e.g., capacity, power gain, and so on)?
	}

	\comment{
		Derivation in the proof of Lemma \ref{lm:pareto} (Appendix \ref{ap:pareto}) is incorrect. Specifically, $(\partial \mathbf{u}^\mathsf{T}) \cdot \mathbf{u}$ is not equal to $\partial (\mathbf{u}^T \mathbf{u})$. Similar error occurs to $\mathbf{v}$. Please note that the derivative of singular value of a matrix with respect to the matrix itself does not exist. In fact, sub-differential should be used instead, e.g., \cite{Cai2010}.
	}

	\comment{
		The rank equation in the proof of Prop. \ref{pp:dof} (Appendix \ref{ap:dof}) seems confusing. The rank of product of two matrices is: $\rank(\mathbf{AB}) = \rank(\mathbf{B}) - \dim(\ker(\mathbf{A}) \cap \ran(\mathbf{B}))$. How could one obtain the rank equation in Appendix \ref{ap:dof}? Please clarify.
	}

	\comment{
		In Sec. \ref{sc:shaping_analytical}, the manuscript provides a number of bounds on singular values of the effective channel $\mathbf{H}$. However, the usefulness of these bounds is unclear. What can these bounds be used for? Note the results Corollaries \ref{co:nd_sv_prod_subset} -- \ref{co:nd_sv_prod_tail} are quite standard results from matrix textbook. The (upper) bound in Corollary \ref{co:nd_power}, which is achieved by aligning its left and right singular-value vectors to that of the forward and backward channels, respectively, is reminiscent of the results in seminal papers on relay beamforming design, e.g., \cite{Rong2009,Rong2009a}.
	}

### 2025-01-03 manuscript/main.tex
:pencil: Update manuscript and add response

% TODO compact or standard SVD

% \title{Channel Shaping Using Beyond Diagonal RIS:\\Analysis, Optimization, and Enhanced Flexibility}
\title{MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS}
% \title{MIMO Channel Shaping Using Beyond Diagonal RIS: Singular Value Redistribution and Rate Maximization}
% \title{Beyond Diagonal RIS: MIMO Channel Shaping and Rate Maximization}
\author{
	\IEEEauthorblockN{
		Yang Zhao, \IEEEmembership{Member, IEEE,}
		Hongyu Li, \IEEEmembership{Graduate Student Member, IEEE,}\\
		Massimo Franceschetti, \IEEEmembership{Fellow, IEEE,}
		and Bruno Clerckx, \IEEEmembership{Fellow, IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

		Massimo Franceschetti is with the Department of Electrical and Computer Engineering, University of California, San Diego, La Jolla CA 92093, USA (e-mail: \href{mailto:massimo@ece.ucsd.edu}{massimo@ece.ucsd.edu}).

@@

\begin{abstract}
	% This paper investigates the extent to which a passive \gls{ris} can redistribute
	% singular values of a point-to-point \gls{mimo} channel for power and rate gains.
	% This paper investigates to what extent can a passive \gls{ris} redistribute
	% singular values to achieve power and rate gains in point-to-point \gls{mimo} channels.
	% the limit to which
	This paper investigates the capability of a passive \gls{ris} to redistribute the singular values of point-to-point \gls{mimo} channels for achieving power and rate gains.
	We depart from the conventional \gls{d}-\gls{ris} with diagonal phase shift matrix
	and adopt a \gls{bd} architecture that offers greater wave manipulation flexibility through element-wise connections.
	% ? not restricted to diagonal scattering matrix and featuring element-wise connections for extra wave manipulation flexibility.
	% featuring element-wise connections for signal amplitude and phase control.
	% featuring element-wise connections for amplitude and phase control.
	% in-group connections between elements, which enables signal amplitude and phase control.
	% This powerful model allows cooperative wave scattering with both amplitude and phase control, and we uniquely attribute the shaping advantage to channel rearrangement and space alignment.
	% This powerful model enables cooperative wave scattering with both amplitude and phase control, and we provide unique shaping insights in terms of channel rearrangement and space alignment.
	% uniquely explain the extra design freedom from off-diagonal entries as channel rearrangement and space alignment.
	% while the shaping advantage from off-diagonal entries is uniquely attributed to channel rearrangement and space alignment.
	% An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} on the Stiefel manifold, which performs multiplicative rotational (instead of add-then-retract) updates for faster convergence.
	% The Pareto frontiers of channel singular values are characterized via a novel geodesic manifold optimization, and the resulting region comprehensively encapsulates relevant metrics (e.g., spectral norm and condition number).
	Specifically, we first
	provide shaping insights by characterizing
	% characterize
	the channel singular value regions attainable by \gls{d}-\gls{ris} and \gls{bd}-\gls{ris} via a novel geodesic optimization.
	Analytical singular value bounds are then derived to explore their shaping limits in typical deployment scenarios.
	% Pareto frontiers of channel singular values via a novel geodesic manifold optimization.
	% channel shaping insights by characterizing the Pareto singular value frontiers via a novel geodesic manifold optimization.
	% The resulting region comprehensively encapsulates most relevant metrics (e.g., spectral norm and condition number).
	% To explore the shaping limits, we also derive
	% 0ol. individual and collective
	% analytical singular value bounds for common \gls{ris} deployment scenarios.
	% in special channel conditions.
	% in two common scenarios: 1) backward/forward channel is rank-deficient and 2) direct channel is blocked.
	% respectively in rank-deficient channels and for fully-connected \gls{bd}-\gls{ris} via matrix analysis.
	% For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% We uniquely attribute the shaping advantage from off-diagonal entries to channel rearrangement and space alignment.
	% Channel rearrangement is a unique feature of off-diagonal entries that allows each group to rearrange and recombine the forward and backward channels.
	% Space alignment generalizes phase matching in \gls{siso} and introduces a trade-off between multiplicative backward-forward and additive direct-indirect channel combinations.
	% Those channel shaping designs have two major implications.
	% First, they demystify the performance gain from off-diagonal entries in terms of {channel rearrangement} and {space alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in channel rearrangement and space alignment.
	% Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} and a shaping-inspired low-complexity approach.
	%  over channel power gain maximization.
	% (i.e., power gain maximization).
	% over channel gain maximization.
	% The former alternates between active and passive beamformers until convergence, while the latter maximizes
	% solves the problem in closed form by a channel power maximization and \gls{mimo} transmission subproblem.
	% shaping-inspired approach.
	Results show that compared to \gls{d}-\gls{ris}, \gls{bd}-\gls{ris} significantly improves the dynamic range of all channel singular values, the trade-off in manipulating them, and thus the channel power and achievable rate.
	Those observations become more pronounced when the number of \gls{ris} elements and \gls{mimo} dimensions increase.
	% Results show that compared to \gls{d}-\gls{ris}, \gls{bd}-\gls{ris} significantly improves the dynamic range of all channel singular values as well as the trade-off in manipulating them.
	% We also notice the power and rate gain of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} increase with the number of scattering elements and \gls{mimo} dimensions.
	Of particular interest, \gls{bd}-\gls{ris} is shown to activate multi-stream transmission at lower transmit power than \gls{d}-\gls{ris}, hence achieving the asymptotic \gls{dof} at low \gls{snr} thanks to its higher flexibility of shaping the distribution of channel singular values.
	% wider dynamic range (with better trade-off) of singular values and significant power and rate gains, especially in large-scale \gls{mimo} systems.

	% indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

	% , and .

	% (i) , and (ii) the gap between the two approaches vanishes as \gls{bd}-\gls{ris} evolves from diagonal to unitary.
	% not only demystify the advantage of \gls{bd}-\gls{ris} in channel rearrangement and space alignment, but also decouple the joint \gls{ris}-transceiver optimization.
	% Those channel shaping designs not only demystify the advantage of \gls{bd}-\gls{ris} in channel rearrangement and space alignment, but also decouple the joint \gls{ris}-transceiver optimization.

	% geodesic \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	% To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in rank-deficient and fully-connected scenarios.
	% As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	% A low-complexity suboptimal solution based on channel shaping is also proposed, where the decoupled problem is formulated as channel power maximization and solved in closed form iteratively.
	% % We now show how channel shaping decouples the design and amounts to a low-complexity suboptimal solution.
	% % A low-complexity suboptimal solution decoupling both blocks is also proposed, where the channel shaping subproblem is formulated as channel power maximization and solved in closed form iteratively.
	% Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions, stemming from stronger channel rearrangement and space alignment capabilities.
\end{abstract}

\begin{IEEEkeywords}
	Reconfigurable intelligent surface, channel singular value redistribution, rate maximization, manifold optimization.
\end{IEEEkeywords}

@@

\begin{section}{Introduction}
	\begin{subsection}{Background}
		Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
		This is empowered by recent advances in \gls{ris}, a programmable passive planar surface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
		A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
		% It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional multi-antenna techniques.
		It enables low-noise full-duplex operation, featuring better flexibility than reflectarrays, lighter footprint than relays, and greater scalability than \gls{mimo} systems.
		One popular \gls{ris} research direction is \emph{joint passive and active beamforming} design with transceivers to enhance a specific performance measure, which has attracted significant interests in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
		While passive beamforming at \gls{ris} suffers attenuation from double fading, it offers better asymptotic behaviors than active beamforming at transceivers (e.g., second-order array gain and fourth-order harvested power \cite{Zhao2022}).
		Another \gls{ris} application is \emph{information modulation} by periodically switching its reflection pattern within the channel coherence time.
		This creates a free-ride message stream with dual benefits: integrating with legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or serving as individual source for low-power uplink communication \cite{Liang2020,Zhao2024,Yang2024}.
		Different from above, \emph{channel shaping} exploits \gls{ris} as a stand-alone device to modify the inherent properties of the wireless environment, for example, compensate for the Doppler effect \cite{Basar2021}, flatten frequency-selective channels \cite{Arslan2022}, improve \gls{mimo} channel rank \cite{Ozdogan2020a}, and artificially diversify channel over time for orthogonal \cite{Yang2019} and non-orthogonal \cite{Chen2023} multiple access schemes.
		This helps to decouple joint beamforming problems into a channel shaping stage and a legacy transceiver design stage, providing a versatile solution for various wireless applications.
		% provides a ubiquitous solution for various wireless applications by decoupling the joint \gls{ris}-transceiver design into a channel design stage and a legacy system design stage.
		% decoupling joint beamforming design by first optimizing the channel and then the transceiver.
		% decouples joint \gls{ris}-transceiver design by first optimizing the channel and then the system, providing a ubiquitous solution for different wireless applications
		% different wireless applications
		% different wireless applications
		% into a channel design stage and a communication system design stage.
		% provides a ubiquitous solution for different wireless applications, decoupling the joint \gls{ris}-transceiver designs into a channel design stage and a communication system design stage.
		% which decouples joint \gls{ris}-transceiver designs by first optimizing the channel and then the communication system.
		% and helps to decouple the \gls{ris}-transceiver design in a distributed manner.
	\end{subsection}

	\begin{subsection}{Related Works}
		At a specific resource block, channel shaping metrics can be categorized into two types.
		% shaping metrics can be classified into two categories:
		\begin{itemize}
			\item \emph{Singular value:} The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}. Those are closely related to performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to minor perturbations of the channel matrix.
			\item \emph{Power:} The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} in point-to-point channels and leakage interference \cite{Santamaria2023a} in interference channels. Those second-order metrics are less informative in \gls{mimo} but easier to analyze and optimize.
		\end{itemize}

		Although above works offered inspiring glimpses into the channel shaping potential of passive \gls{ris}, none attempted to disclose the entire attainable channel singular value region.
		% to characterize the channel singular value region manipulated by passive \gls{ris}.
		% no previous work has characterized the \gls{mimo} channel singular value region manipulated by any type of \gls{ris}.
		Most relevant literature \cite{ElMossallamy2021,Meng2023,Zheng2022,Huang2023,Bafghi2022,Zheng2023,Chae2023,Wu2019,Santamaria2023a} have also been limited to a \gls{d}-\gls{ris} model where each element is connected to a dedicated impedance and disconnected from others.
		As such, wave impinging on one element is entirely scattered by the same element.
		% is tuned by a dedicated impedance and acts as an \emph{individual} scatterer \cite{Wu2020}.
		This simple architecture is modeled by a diagonal scattering matrix with unit-magnitude diagonal entries, which only applies a phase shift to the incoming signal.
		The idea was soon generalized to \gls{bd}-\gls{ris} with group-connected architecture \cite{Shen2020a}, where adjacent elements within the same group are connected via passive reconfigurable components\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices, respectively.}.
		This allows wave impinging on one element to propagate within the circuit and depart partially from any element in the same group.
		It can thus manipulate both amplitude and phase of the scattered wave while remaining passive.
		Such a powerful model can be realized at reduced hardware cost using tree- and forest-connected architectures by graph theory \cite{Nerini2024}.
		\gls{bd}-\gls{ris}
		% also supports
		can also function in
		% hybrid transmitting-and-reflecting mode \cite{Li2023b} and
		multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
		Practical challenges such as channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a} have also been studied in recent literature.
		Its beamforming superiority over \gls{d}-\gls{ris} has been proved in \gls{siso} and \gls{miso} systems \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023,Fang2023,Zhou2023,Li2023c,Soleymani2024}, however, the interplay between \gls{bd}-\gls{ris} and \gls{mimo} is still in the infancy stage.
		% On the other hand, the beamforming superiority of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} has been extensively studied in \gls{siso} and \gls{miso} communication systems \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023,Fang2023,Zhou2023,Li2023c,Soleymani2024}.
		% % , where the problems have been cast as single-user \gls{snr} maximization \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and multi-user \gls{wsr} maximization \cite{Fang2023,Zhou2023,Li2023c,Soleymani2024}.
		% However, the interplay between \gls{bd}-\gls{ris} and \gls{mimo} is still in the infancy stage.
		The authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design for a fully-connected \gls{bd}-\gls{ris}-aided \gls{mimo} system where the direct channel is negligible.
		% 		The work
		% 		% For a massive \gls{mimo} without direct link,
		% 		Similar assumptions were also adopted in
		% [47], which introduces a transmitter-side BD-RIS to massive
		% MIMO that exploits statistical Channel State Information (CSI)
		% for improved spectral efficiency
		% 		% with negligible direct link and \gls{bd}-\gls{ris}.
		% 		% Those assumptions simplify the problem but limit the generality of the results.
		A transmitter-side \gls{bd}-\gls{ris} was introduced to
		% improve the spectral efficiency of massive \gls{mimo} systems where the direct channel is negligible \cite{Mishra2024}.
		massive \gls{mimo} systems that exploits statistical \gls{csi} for improved spectral efficiency \cite{Mishra2024}, which again assumed negligible direct channel and fully-connected \gls{bd}-\gls{ris}.
		Received power maximization with continuous-valued and discrete-valued \gls{bd}-\gls{ris} have been tackled respectively in closed form \cite{Nerini2023} and by machine learning approach \cite{Nerini2023b}, but the corresponding single-stream transceiver is rate-suboptimal.
		%  and the \gls{bd}-\gls{ris} design problem is equivalent to \gls{siso} shaping.

		% in single-stream \gls{mimo} transmission

		% in the literature, but the resulting single-stream transceiver is rate-suboptimal and the corresponding \gls{bd}-\gls{ris} design is equivalent to \gls{siso} shaping.
		% A practical frequency-dependent \gls{bd}-\gls{ris} model has been recently proposed for multi-band multi-cell \gls{mimo} networks to facilitate practical deployments \cite{Sena2024}.
		% Although the results are promising, the channel shaping capability from off-diagonal entries deserves further investigation, and there lacks a universal yet efficient \gls{bd}-\gls{ris} design framework.
	\end{subsection}

	\begin{subsection}{Contributions}
		% We notice that there lacks a comprehensive study of \gls{bd}-\gls{ris} in point-to-point \gls{mimo} systems

		% This paper investigates the shaping capability of passive \gls{ris} in point-to-point \gls{mimo} channels.
		% We aim to answer the ultimate question:
		% \emph{What is the channel singular value (and power gain) shaping capability of \gls{bd}-\gls{ris}?}

		This paper is motivated by a fundamental question:
		\emph{What is the singular value (and power gain) shaping capability of a passive \gls{ris} in point-to-point \gls{mimo} channels?}
		We aim for a comprehensive answer via analysis and optimization.
		The contributions are summarized below.

		% This paper aims to answer a critical question:
		% \emph{What is the singular value (and power gain) shaping capability of passive \gls{ris} in point-to-point \gls{mimo} channels?}
		% We provide a comprehensive answer through theoretical analysis and numerical optimization.

		% This paper investigates the shaping capability of passive \gls{ris} in point-to-point \gls{mimo} channels.
		% We aim to answer the ultimate question:
		% \emph{What is the channel singular value and power gain shaping capability of \gls{bd}-\gls{ris}?}
		% \emph{What is the ultimate channel singular value shaping capability of a passive \gls{ris}?}

>>>>

% \title{Channel Shaping Using Beyond Diagonal RIS:\\Analysis, Optimization, and Enhanced Flexibility}
\title{MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS}
% \title{MIMO Channel Shaping Using Beyond Diagonal RIS: Singular Value Redistribution and Rate Maximization}
% \title{Beyond Diagonal RIS: MIMO Channel Shaping and Rate Maximization}
\author{
	\IEEEauthorblockN{
		Yang Zhao, \IEEEmembership{Member, IEEE,}
		Hongyu Li, \IEEEmembership{Graduate Student Member, IEEE,}\\
		Massimo Franceschetti, \IEEEmembership{Fellow, IEEE,}
		and Bruno Clerckx, \IEEEmembership{Fellow, IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K.
		% (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

		Massimo Franceschetti is with the Department of Electrical and Computer Engineering, University of California, San Diego, La Jolla CA 92093, USA.
		% (e-mail: \href{mailto:massimo@ece.ucsd.edu}{massimo@ece.ucsd.edu}).

@@

\begin{abstract}
	This paper investigates the limits to which a passive \gls{ris} can redistribute the singular values of a point-to-point \gls{mimo} channel for providing power and rate gains.
	We depart from conventional \gls{d}-\gls{ris} with diagonal phase shift matrix
	and adopt a \gls{bd} architecture featuring group-wise connections for superior branch matching and mode alignment capabilities.
	% wave manipulation flexibility.
	Specifically, analytical channel singular value bounds are derived under the scenario of rank-deficient backward/forward channel or negligible direct channel, which imply the \gls{bd}-\gls{ris} shaping advantage in terms of dynamic range and \gls{dof}.
	A numerical method is then proposed to optimize the \gls{bd}-\gls{ris} for a special class of singular value functions and invoked in a Pareto problem to reveal the singular value region.
	As a side product, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} and a low-complexity shaping-inspired approach, and extend both to \gls{wsr} maximization in \gls{mimo} interference channels.
	Results show that compared to \gls{d}-\gls{ris}, \gls{bd}-\gls{ris} significantly improves the dynamic range of all channel singular values and the trade-off in manipulating them, resulting in enhanced power gain and achievable rate.
	Those gaps become more pronounced when the number of \gls{ris} elements or \gls{mimo} dimensions increase.
	Of particular interest, \gls{bd}-\gls{ris} is shown to activate multi-stream transmission (hence achieving the asymptotic \gls{dof}) at much lower transmit power than \gls{d}-\gls{ris} thanks to its singular value redistribution capability.
\end{abstract}

\begin{IEEEkeywords}
	MIMO, RIS, channel shaping, rate maximization, singular value bounds, manifold optimization, interference channel.
\end{IEEEkeywords}

@@

\begin{section}{Introduction}
	\begin{subsection}{Background}
		Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
		This is empowered by recent advances in \gls{ris}, a programmable passive planar surface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
		A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
		It enables low-noise full-duplex operation, featuring better flexibility than reflectarrays, lighter footprint than relays, and greater scalability than \gls{mimo} systems.
		One popular \gls{ris} research direction is \emph{joint passive and active beamforming} design with transceivers to enhance a specific performance measure, which has attracted significant interests in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
		While passive beamforming at \gls{ris} suffers attenuation from double fading, it offers better asymptotic behaviors than active beamforming at transceivers (e.g., second-order array gain and fourth-order harvested power \cite{Zhao2022}).
		Another \gls{ris} application is \emph{information modulation} by periodically switching its reflection pattern within the channel coherence time.
		This creates a free-ride message stream with dual benefits: integrating with legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or serving as individual source for low-power uplink communication \cite{Liang2020,Zhao2024,Yang2024}.
		Different from above, \emph{channel shaping} exploits \gls{ris} as a stand-alone device to modify the inherent properties of the wireless environment, for example, compensate for the Doppler effect \cite{Basar2021}, flatten frequency-selective channels \cite{Arslan2022}, improve \gls{mimo} channel rank \cite{Ozdogan2020a}, and artificially diversify channel over time for orthogonal \cite{Yang2019} and non-orthogonal \cite{Chen2023} multiple access schemes.
		This helps to decouple joint beamforming problems into a channel shaping stage and a legacy transceiver design stage, providing a versatile solution for various wireless applications.
	\end{subsection}

	\begin{subsection}{Related Works}
		At a specific resource block, channel shaping metrics can be classified into two categories.
		\begin{itemize}
			\item \emph{Singular value:} The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}. Those are closely related to performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to minor perturbations of the channel matrix.
			\item \emph{Power:} The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} in point-to-point channels and leakage interference \cite{Santamaria2023a} in interference channels. Those second-order metrics are less informative in \gls{mimo} but easier to analyze and optimize.
		\end{itemize}

		Although above works offered inspiring glimpses into the channel shaping potential of passive \gls{ris}, none attempted to disclose the entire singular value region.
		Most relevant literature \cite{ElMossallamy2021,Meng2023,Zheng2022,Huang2023,Bafghi2022,Zheng2023,Chae2023,Wu2019,Santamaria2023a} have also been limited to a \gls{d}-\gls{ris} model where each element is connected to a dedicated impedance and disconnected from others.
		As such, wave impinging on one element is entirely scattered by the same element.
		This simple architecture is modeled by a diagonal scattering matrix with unit-magnitude diagonal entries, which only applies a phase shift to the incoming signal.
		The idea was soon generalized to \gls{bd}-\gls{ris} with group-connected architecture \cite{Shen2020a}, where adjacent elements within the same group are connected via passive reconfigurable components\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices, respectively.}.
		This allows wave impinging on one element to propagate within the circuit and depart partially from any element in the same group.
		It can thus manipulate both amplitude and phase of the scattered wave while remaining passive.
		Such a powerful model can be realized at reduced hardware cost using tree- and forest-connected architectures by graph theory \cite{Nerini2024}.
		\gls{bd}-\gls{ris} can also function in
		% hybrid transmitting-and-reflecting mode \cite{Li2023b} and
		multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
		Practical challenges such as channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a} have also been studied in recent literature.
		Its beamforming effectiveness over \gls{d}-\gls{ris} and energy efficiency over active \gls{ris} and relay systems have been proved in \gls{siso} and \gls{miso} systems \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023,Fang2023,Zhou2023,Li2023c,Soleymani2024}.
		However, the interplay between \gls{bd}-\gls{ris} and \gls{mimo} is still in the infancy stage.
		The authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design for a fully-connected \gls{bd}-\gls{ris}-aided \gls{mimo} system where the direct channel is negligible.
		A transmitter-side \gls{bd}-\gls{ris} was introduced to
		massive \gls{mimo} systems that exploits statistical \gls{csi} for improved spectral efficiency \cite{Mishra2024}, which again assumed negligible direct channel and fully-connected \gls{bd}-\gls{ris}.
		Received power maximization with continuous-valued and discrete-valued \gls{bd}-\gls{ris} have been tackled respectively in closed form \cite{Nerini2023} and by machine learning approach \cite{Nerini2023b}, but the corresponding single-stream transceiver is rate-suboptimal.
	\end{subsection}

	\begin{subsection}{Contributions}
		\label{sc:contributions}
		This paper is motivated by a fundamental question:
		\emph{What is the singular value shaping capability of a passive \gls{ris} in \gls{mimo} channels?}
		We aim for a comprehensive answer via analysis and optimization.
		The contributions are summarized below.

		First, we pioneer \gls{bd}-\gls{ris} study in general \gls{mimo} channels and interpret its shaping potential as branch matching and mode alignment.
		Branch matching refers to pairing and combining the branches (i.e., entries) of backward and forward channels associated with each \gls{bd}-\gls{ris} group.
		Mode alignment refers to aligning and ordering the modes (i.e., singular vectors) of indirect channels with those of direct channel.
		The former origins uniquely from the off-diagonal entries of the scattering matrix of \gls{bd}-\gls{ris}.

		Second, we provide an analytical answer to the shaping question under specific channel conditions.
		It is proved that \gls{bd}-\gls{ris} may achieve a larger or smaller number of \gls{mimo} \gls{dof} than \gls{d}-\gls{ris}.
		When the backward or forward channel is rank-deficient, we derive asymptotic singular value bounds applying to both \gls{d}- and \gls{bd}-\gls{ris}.
		When the direct channel is negligible, we recast the shaping question for fully-connected \gls{bd}-\gls{ris} as a well-studied linear algebra question and provide tight bounds on channel singular values, power gain, and capacity.
		The corresponding \gls{ris} scattering matrices are also derived in closed form.

		Third, we provide a numerical answer to the shaping question by exploiting a geodesic \gls{rcg} method to optimize the \gls{bd}-\gls{ris} for a special class of singular value functions.
		It compares favorably to legacy \gls{ris} designs in that the updates are performed multiplicatively along the shortest paths on the manifold for accelerated convergence.
		The method is then invoked in a Pareto problem to reveal the entire singular value region, which generalizes most relevant metrics and provides an intuitive shaping benchmark.

		Fourth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} and a low-complexity shaping-inspired approach.
		The former iteratively updates the passive beamforming by geodesic \gls{rcg} and the active beamforming by eigenmode transmission, until convergence.
		The latter exploits the \gls{bd}-\gls{ris} to shape the channel once for maximum power gain and then performs eigenmode transmission.
		% Interestingly, the rate deficit from the low-complexity design diminishes as the \gls{bd}-\gls{ris} evolves towards fully-connected.
		We also extend both approaches to \gls{wsr} maximization in \gls{mimo} interference channels where the low-complexity design aims for minimum leakage interference.

		Fifth, we conduct extensive simulations to validate the analytical bounds and numerical methods.
		It is concluded that:
		\begin{enumerate}
			\item \gls{bd}-\gls{ris} significantly improves the dynamic range of all channel singular values and the trade-off in manipulating them, resulting in enhanced power gain and achievable rate;
			\item The shaping advantages of \gls{bd}-\gls{ris} increase with the number of scattering elements and \gls{mimo} dimensions;
			\item \gls{bd}-\gls{ris} can activate multi-stream transmission (hence achieving the asymptotic \gls{dof}) at low transmit power thanks to its superior singular value redistribution capability;
			\item The rate gap between the optimal and low-complexity beamforming designs diminishes as the \gls{ris} evolves from \gls{d} to fully-connected \gls{bd};
			\item The additional optimization cost of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} is affordable and the geodesic \gls{rcg} method is efficient on large-scale problems;
			\item All proposed asymmetric \gls{bd}-\gls{ris} designs can be modified for symmetry with reasonable performance cost.
		\end{enumerate}
	\end{subsection}
